---
title: "Assignment 1 ML"
author: "Ashley Riley"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**This assignment will concentrate on using R and Git. Specifically, you will do the following:**

1. Download a dataset from the web. You may use any source, but specify the source in
your code. Also ensure that the data has a mix of quantitative and qualitative
(categorical) variables.

Data source: Baseball-Reference.com - 2023 hitting statistics


2. Import the dataset into R

```{r}
##import data from .csv file##
df<-read.csv("/Users/ashleyriley/Desktop/Fundamentals of Machine Learning/RStudio/2023 hitting stats.csv" , header=TRUE, sep = ",")
head(df, 10)##there is a lot of data so limited to top 10 for display purposes##
```
3. Print out descriptive statistics for a selection of quantitative and categorical variables.
```{r}
##count of the number of players on each team##
table(df$Tm)
```
```{r}
##count of hits by postition type##
table(df$Pos..a0Summary)
```
```{r}
##avg age of players in 2023##
mean(df$Age, na.rm=TRUE)
```

4. Transform at least one variable. It doesn't matter what the transformation is.
```{r}
##changing the column name that presented oddly from the source data into something more human readable##
cn= colnames(df)
cn[cn=="Pos..a0Summary"] = "Position Type"
colnames(df) = cn
head(df)##displaying header names to show changed value##
```
5. Plot at least one quantitative variable, and one scatterplot
```{r}
##df$Age = as.integer(as.character(df$Age))
##print(sapply(df,class))
##x<-df$Age
##y<-df$OBP
##plot(x, y, main = "Age Impact to On-Base Percentage", xlab = "Player Age", ylab = "On-base Percentage (OBP)", pch=19, frame = FALSE)
##abline(lm(y-x, data = df), col = "blue")

##scatter plot of age vs. on-base % (OBP)##
input<-df[,c('Age','OBP')]
plot(x = input$Age, y = input$OBP, xlab = "Player Age", ylab = "On-base Percentage (OBP)", xlim = c(20,40), ylim = c(0,1), main = "Age vs. OBP")
```
```{r}
##Plotting quantitative variable##
tapply(df$Age, df$Tm, mean, na.rm = TRUE)

boxplot(df$Age~df$Tm, col = 'green', main = 'Avg Age by Team', xlab = 'Team', ylab = 'Age')



```




6. Upload your R program, and any associated datafiles to your git account. Remember to
create a separate repository for this class.
7. Paste the address to your repository in the assignment submission box here in Canvas.
